19Q2

### 2019-05-17 19Q2.Rmd
updated Rmd for merging drug data

This comprehensive cancer cell line information was curated by both Daniel Charytonowicz and myself.
```{r dan_ccl_database}
ccl_converter <- read.delim("./data_munging/cell_line_database_v1_20180911.tsv", row.names = 1, sep = "\t", header = TRUE)
colnames(ccl_converter)[colnames(ccl_converter) == "Broad_ID"] <- "DepMap_ID"
ccl_lineage_converter <- unique(ccl_converter[, c("DepMap_ID", "grouped_general_doid", "group_general_lineage_name")])
lineage_converter <- unique(select(ccl_lineage_converter, grouped_general_doid, group_general_lineage_name))
ccl_histologies <- read.delim("./data_munging/ccl_histologies_metadata.csv", sep = ",", header = TRUE, row.names = 1)
```

@@

g2p_dgl <-  as.character(unique(g2p$Drug_Gene_Lin[!is.na(g2p$Drug_Gene_Lin)]))
g2p_dg <- as.character(unique(g2p$Drug_Gene[!is.na(g2p$Drug_Gene)]))
g2p_gl <- as.character(unique(g2p$Drug_Gene[!is.na(g2p$Gene_Lin)]))
g2p_drugs <- unique(sapply(strsplit(g2p_dgl, "_", fixed = TRUE), function(x) x[1]))
g2p_genes <- unique(sapply(strsplit(g2p_dgl, "_", fixed = TRUE), function(x) x[2]))
g2p_lins <- unique(sapply(strsplit(g2p_dgl, "_", fixed = TRUE), function(x) x[3]))
```

@@

```{r omics_functions}
MergeOmicsG2P <- function(database, maf_df, ge_df, cn_df, outfile) {
  # Merge mutation status, CN, and GE omics data with drugscreen results and metadata
  if(database %in% c("CCLE", "CTRP", "GDSC")) {
    filt <- auc %>% filter(Database == database) %>% select(Cell_line, Histology, TCGA_classification, CCLE_Name, DepMap_ID, Drug, AUC_IC50)
    print(paste0("Processing ", database, " data."))
    colnames(filt)[colnames(filt) == "AUC_IC50"] <- "Score"

    # Get all G2P gene-cell line combinations
    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = g2p_genes)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, DepMap_ID, Primary.Disease), by = c("Cell_line", "Histology", "CCLE_Name", "DepMap_ID"), all.y = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_druggenes, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgls, "Yes", "No")
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "DepMap_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Get all G2P gene-cell line combinations
    data <-  filter(data, Gene %in% g2p_genes)
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, DepMap_ID, Primary.Disease), by = "DepMap_ID", all.y = TRUE)
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
    data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  }

  ## Summarize number of mutant and Wildtype cell lines
  data_muts_summ <- data %>% group_by(Gene) %>% summarize(N_Wildtype = length(Mutation_Status) - sum(Mutation_Status == 1), N_Mutant = sum(Mutation_Status == 1))
  ## Merge test results back into full dataset, which restores information lost in the summarization
  data <- merge(data_muts_summ, data, by = "Gene")
  print("Merged mutation calls.")
  # Merge GE data
  data <- merge(data, ge_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
  print("Merged gene expression data.")
  # Merge CN data
  data <- merge(data, cn_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
  print("Merged copy number data.")
  # Drop cell lines not in DepMap
  data <- filter(data, !is.na(DepMap_ID) & !(DepMap_ID %in% c("ACH-001182", "ACH-001451")))
  # Save and return dataframe
  data <- unfactorize(data)
  saveRDS(data, outfile, compress = "xz")
  return(data)
}
```

@@

ccle <- MergeOmicsG2P(database = "CCLE", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/ccle_data_19Q2.rds")
# ctrp <- MergeOmicsG2P(database = "CTRP", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./../htscreens_giant_files/ctrp_data_19Q2.rds")
# gdsc <- MergeOmicsG2P(database = "GDSC", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, outfile = "./data_munging/rds/gdsc_data_19Q2.rds")
```

```{r}
ccle_missense <- filter(ccle, Variant_Classification == "Missense_Mutation")
ccle_missense_grid <- expand.grid("Gene" = unique(ccle_missense$Gene), "DepMap_ID" = unique(ccle_missense$DepMap_ID))
ccle_missense_grid$Variant_Classification <- NA
ccle_missense <- merge(ccle_missense, ccle_missense_grid, by = c("Gene", "DepMap_ID", "Variant_Classification"), all = TRUE)
```

```{r load_drug_rds, eval=TRUE}
ccle_ptmuts <- readRDS("./data_munging/rds/ccle_data_19Q2_g2pgenes_ptmuts.rds") %>% drop_na(Drug)
ctrp_ptmuts <- readRDS("./../htscreens_giant_files/ctrp_data_19Q2_g2pgenes_ptmuts.rds") %>% drop_na(Drug)
gdsc_ptmuts <- readRDS("./data_munging/rds/gdsc_data_19Q2_g2pgenes_ptmuts.rds") %>% drop_na(Drug)

>>>>

This comprehensive cancer cell line information was curated by both Daniel Charytonowicz and myself.
```{r dan_ccl_database}
ccl_converter <- read.delim("./data_munging/cell_line_database_v1_20180911.tsv", row.names = 1, sep = "\t", header = TRUE)
colnames(ccl_converter)[colnames(ccl_converter) == "Broad_ID"] <- "DepMap_ID"
ccl_lineage_converter <- unique(ccl_converter[, c("DepMap_ID", "grouped_general_doid", "group_general_lineage_name")])
lineage_converter <- unique(select(ccl_lineage_converter, grouped_general_doid, group_general_lineage_name))
ccl_histologies <- read.delim("./data_munging/ccl_histologies_metadata.csv", sep = ",", header = TRUE, row.names = 1)
colnames(ccl_histologies) <- c("Cell_line", "CCLE_histology", "TCGA_classification", "CCLE_name", "DepMap_ID")
```

@@

g2p_dgl <-  as.character(unique(g2p$Drug_Gene_Lin[!is.na(g2p$Drug_Gene_Lin)]))
g2p_dg <- as.character(unique(g2p$Drug_Gene[!is.na(g2p$Drug_Gene)]))
g2p_gl <- as.character(unique(g2p$Gene_Lin[!is.na(g2p$Gene_Lin)]))
g2p_drugs <- as.character(unique(g2p$Drug[!is.na(g2p$Drug)]))
g2p_genes <- as.character(unique(g2p$Gene[!is.na(g2p$Gene)]))
g2p_lins <- as.character(unique(g2p$group_general_lineage_name[!is.na(g2p$group_general_lineage_name)]))
```

@@

```{r omics_functions}
MergeOmicsG2P <- function(database, maf_df, ge_df, cn_df, gene_list, outfile) {
  # Merge mutation status, CN, and GE omics data with drugscreen results and metadata
  if(database %in% c("CCLE", "CTRP", "GDSC")) {
    filt <- auc %>% filter(Database == database) %>% select(Cell_line, CCLE_histology, TCGA_classification, CCLE_name, DepMap_ID, Drug, AUC_IC50)
    print(paste0("Processing ", database, " data."))
    colnames(filt)[colnames(filt) == "AUC_IC50"] <- "Score"

    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = gene_list)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, CCLE_histology, CCLE_name, DepMap_ID, disease), by = c("Cell_line", "CCLE_histology", "CCLE_name", "DepMap_ID"), all.y = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_dg, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgl, "Yes", "No")
    data$Gene_Lin <- paste0(data$Gene, "_", data$Gene, "_", data$disease)
    data$G2P_Gene_Lin <- ifelse(data$Gene_Lin %in% g2p_gl, "Yes", "No")
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
    data$Mutation <- ifelse(is.na(data$Mutation), 0, data$Mutation)
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "DepMap_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, CCLE_histology, CCLE_name, DepMap_ID, disease), by = "DepMap_ID", all.y = TRUE)
    print("Merged metadata.")
    # Merge mutation calls
    data <- unfactorize(data)
    data <- merge(data, maf_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
    data$Mutation <- ifelse(is.na(data$Mutation), 0, data$Mutation)
  }

  ## Summarize number of mutant and Wildtype cell lines
  data_muts_summ <- data %>% group_by(Gene) %>% summarize(N_Wildtype = length(Mutation) - sum(Mutation == 1), N_Mutant = sum(Mutation == 1))
  ## Merge test results back into full dataset, which restores information lost in the summarization
  data <- merge(data_muts_summ, data, by = "Gene")
  print("Merged mutation calls.")
  # Merge GE data
  data <- merge(data, ge_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
  print("Merged gene expression data.")
  # Merge CN data
  data <- merge(data, cn_df, by = c("Gene", "DepMap_ID"), all.x = TRUE)
  print("Merged copy number data.")
  # Drop cell lines not in DepMap
  data <- filter(data, !is.na(DepMap_ID) & !(DepMap_ID %in% c("ACH-001182", "ACH-001451")))
  # Save and return dataframe
  data <- unfactorize(data)
  saveRDS(data, outfile, compress = "xz")
  return(data)
}
```

@@

ccle <- MergeOmicsG2P(database = "CCLE", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, gene_list = g2p_genes, outfile = "./../htscreens_giant_files/ccle_data_19Q2.rds")
ctrp <- MergeOmicsG2P(database = "CTRP", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, gene_list = g2p_genes, outfile = "./../htscreens_giant_files/ctrp_data_19Q2.rds")
gdsc <- MergeOmicsG2P(database = "GDSC", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, gene_list = g2p_genes, outfile = "./../htscreens_giant_files/gdsc_data_19Q2.rds")

# all_genes <- intersect(unique(maf_df$Gene), intersect(unique(cn_melt$Gene), unique(ge_melt$Gene)))
# ccle <- MergeOmicsG2P(database = "CCLE", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, gene_list = all_genes, outfile = "./../htscreens_giant_files/ccle_data_19Q2.rds")
# ctrp <- MergeOmicsG2P(database = "CTRP", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, gene_list = all_genes, outfile = "./../htscreens_giant_files/ctrp_data_19Q2.rds")
# gdsc <- MergeOmicsG2P(database = "GDSC", maf_df = maf_df, ge_df = ge_melt, cn_df = cn_melt, gene_list = all_genes, outfile = "./../htscreens_giant_files/gdsc_data_19Q2.rds")
```

```{r load_drug_rds, eval=FALSE}
ccle <- readRDS("./data_munging/rds/ccle_data_19Q2_g2pgenes_ptmuts.rds") %>% drop_na(Drug)
ctrp <- readRDS("./data_munging/rds/ctrp_data_19Q2_g2pgenes_ptmuts.rds") %>% drop_na(Drug)
gdsc <- readRDS("./data_munging/rds/gdsc_data_19Q2_g2pgenes_ptmuts.rds") %>% drop_na(Drug)
